---
title: "Quantifying the number of *rrn* operons across taxonomic ranks"
author: "G Bhatti; P Schloss"
date: "9/13/2024"
output:
 github_document:
   html_preview: false
editor_options: 
  chunk_output_type: console
---

```{r setup, message=FALSE}
library(tidyverse)
library(here)
library(ggridges)

metadata<- read_tsv(here("data/references/genome_id_taxonomy.tsv"),
                    col_types = cols(.default = col_character())) |> 
  mutate(strain=if_else(scientific_name==species,NA_character_,scientific_name)) |> 
  select(-scientific_name)


asv<- read_tsv(here("data/processed/rrnDB.count_tibble"),
               col_types = cols(.default = col_character(),
                                count= col_integer()))


metadata_asv<- inner_join(metadata, asv, by=c("genome_id"="genome"))
```

### Plot the number of *rrn* copies per taxonomic rank

Our analysis will use fill length sequences. We want to count and plot the number
of copies per *taxonomic rank*. Before calculating the averages for each taxonomic 
group, we should calculate average number of copies for each species. This will 
allow us to control for uneven number of genomes in each species.
within each rank.

```{r}
rank_taxon_rrns<- metadata_asv |> 
  filter(region =="v19") |> 
  group_by(kingdom, phylum, class, order, family, genus, species, genome_id) |> 
  summarize(n_rrns=sum(count),.groups = "drop") |> 
  group_by(kingdom, phylum, class, order, family, genus, species) |> 
  summarise(mean_rrns= mean(n_rrns),.groups = "drop") |> 
  pivot_longer(-mean_rrns,
               names_to = "rank",
               values_to="taxon") |> 
  drop_na(taxon) |> 
  mutate(rank=factor(rank,
                     levels=c("kingdom","phylum","class","order",
                              "family","genus","species","strain"))) |> 
  group_by(rank,taxon) |> 
  summarise(mean_rrns=mean(mean_rrns),.groups = "drop")

mean_of_means<- rank_taxon_rrns |> 
  group_by(rank) |> 
  summarise(mean_of_means=mean(mean_rrns),.groups = "drop")

median_of_means<- rank_taxon_rrns |> 
  group_by(rank) |> 
  summarise(median_of_means=median(mean_rrns),.groups = "drop")

```



```{r}
jitter_width <- 0.3
n_ranks<-nrow(mean_of_means)
rank_taxon_rrns |> 
  ggplot(aes(x=rank,y=mean_rrns)) +   

  geom_jitter(width=jitter_width,color="gray") +
  # geom_point(data=mean_of_means,aes(x=rank,y=mean_of_means),
  #            color="red",size=3,inherit.aes = F)+
  # geom_line(data=mean_of_means,aes(x=rank,y=mean_of_means),
  #           color="red",size=2,group=1,
  #           inherit.aes = F)+
  
  geom_segment(data=mean_of_means,
               aes(x=1:n_ranks-jitter_width,xend=1:n_ranks+jitter_width,
                   y=mean_of_means,yend= mean_of_means),
               color="red",size=2,inherit.aes = F,lineend = "round")+
 # geom_boxplot(outlier.shape = NA,fill=NA)+
  theme_classic()+
  labs(x=NULL,
       y="Mean number of rrn copies per genome",
       title= "There is wide variation at the species level in the number of rrn copies",
       subtitle = "Each point represents a single taxon within that rank, numbers based\non average species copy number")+
  scale_x_discrete(breaks=c("kingdom","phylum","class","order",
                            "family","genus","species"),
                   labels=str_to_title(c("kingdom","phylum","class","order",
                            "family","genus","species")))
```

Here's another way of looking at the data...

```{r}
rank_taxon_rrns |> 
  ggplot(aes(x=mean_rrns,y=rank)) +  
  geom_density_ridges(stat="binline",binwidth=1,scale=0.9)+
  # geom_density(alpha=0.2,adjust=1.5)
  # geom_histogram(binwidth =1,position = "dodge")
  geom_point(data=median_of_means,aes(y=1:n_ranks+0.1,x=median_of_means),
             color="red",size=3,inherit.aes = F)+

   theme_classic()+
  labs(y=NULL,
       x="Mean number of rrn copies per genome",
       title= "The distribution of rrn copies per genome is consistent across ranks",
       subtitle = "Each point represents a single taxon within that rank, numbers based\non average species copy number")+
  scale_y_discrete(breaks=c("kingdom","phylum","class","order",
                            "family","genus","species"),
                   labels=str_to_title(c("kingdom","phylum","class","order",
                            "family","genus","species")))
  
```

* Bacteria have more copies than Archaea
* Even after correcting for the number genomes per species, there is is wide variation in the number of *rrn* operons per taxonomic group

